Copy number inference with Ewing sarcoma samples

### 2024-05-03 analyses/ewings-cell-type/03-cnv-inference.Rmd
compare ref to no ref

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")

# source in helper functions
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

# output files 
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)
```

## Analysis content

### CopyKAT

```{r}
ck_results <- copykat::copykat(rawmat = as.matrix(counts(sce)),
                               id.type = "E", 
                               n.cores = 4)
```

Save these results so we don't need to re-run! 
```{r}
table(ck_results$prediction$copykat.pred)
# temp save ck_results 
readr::write_rds(ck_results, "results/SCPCL000490_copykat_results.rds")
```

```{r}
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      )

ck_preds <- ck_results$prediction |> 
  dplyr::rename(
    barcodes = cell.names
  )

cnv_df <- umap_df |> 
  dplyr::left_join(ck_preds)
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = copykat.pred)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw()
```

```{r}
# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)

cnv_df <- cnv_df |> 
  dplyr::left_join(manual_classifications_df)
```

```{r}
caret_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  ))

caret::confusionMatrix(table(caret_df$marker_gene_classification, 
      caret_df$copykat))
```

Look at how many cells are aneuploid vs. diploid
Compare tumor vs. normal from manual to copykat using confusion matrix

Try and repeat but with providing a baseline reference

>>>>

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
```

```{r}
# source in helper functions for make_jaccard_matrix() and jaccard()
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

# copykat results 
results_dir <- file.path(module_base, "results", "copykat", params$library_id)
results_filenames <- c(
  no_ref = glue::glue("{params$library_id}_no_normal_copykat_prediction.txt"), 
  with_ref = glue::glue("{params$library_id}_with_normal_copykat_prediction.txt")
) 
ck_results_files <- file.path(results_dir, results_filenames) |> 
  purrr::set_names(names(results_filenames))
```

```{r}
# read in processed sce
sce <- readr::read_rds(sce_file)

# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)

# read in ck results 
ck_results_df <- ck_results_files |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

## CopyKAT results

```{r}
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      )

cnv_df <- umap_df |> 
  # first add manual annotations
  dplyr::left_join(manual_classifications_df) |> 
  # now add copykat results
  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names"))
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = copykat.pred)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  facet_wrap(vars(reference_used))
```

```{r}
caret_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  )) |> 
  split(cnv_df$reference_used)

caret_df |> 
  purrr::imap(\(df, ref_type){
    glue::glue("\n\n================== {ref_type} ====================")
    caret::confusionMatrix(
      table(
        df$marker_gene_classification, 
        df$copykat)
    )
  })
```

```{r}
# calculate Jaccard similarity index for each reference type
ref_types <- c("no_ref", "with_ref")
jaccard_matrices <- ref_types |>
  purrr::map(\(ref_type) {
    jaccard_df <- cnv_df |> 
      dplyr::filter(reference_used == ref_type)

    make_jaccard_matrix(
      cnv_df,
      "marker_gene_classification",
      "copykat.pred"
    )
  }) |> 
  purrr::set_names(ref_types)
```

```{r}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

# heatmaps comparing tumor/normal annotations manually vs. copyKAT
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(jaccard_mtx, ref_type) {
      ComplexHeatmap::Heatmap(
        t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = ref_type,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = ref_type == "no_ref",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

Now compare to CellAssign and SingleR annotations.

```{r}
celltype_columns <- c(
  "singler_celltype_annotation",
  "cellassign_celltype_annotation"
)

# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      cnv_df,
      "copykat.pred",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```

```{r, fig.height=10, fig.width=7}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
